\chapter{CONCLUSION}
There is an uncomfortable tension between computer forensics and privacy. Computer forensics draws attention to all of the pieces of
information an individual leaves behind on a device.  Privacy preserving technologies strive to hide, destroy, or otherwise prevent
the recovery of these same bits of information. A delicate balance between the two must be struck, as individuals may find
themselves requiring the use of either forensic tools or privacy tools, depending on the situation.  Android forensics is developing
at an astonishing pace.  Already, live-system forensics is being introduced for Android devices, and soon live memory acquisition of
phones and tablets will be commonplace. For activists, dissidents, and plain-old citizens crossing borders, it is imperative that
methods exist to protect mobile data against these increasingly sophisticated acquisition techniques.  

This paper gave a short overview of Android forensics techniques, emphasizing the methods by which the data on an Android device may
be acquired. These methods explicitly attempt to subvert the privacy of data on a device by making a low-level copy, and therefore
serve as an excellent litmus test for privacy tools. Encrypting a device, so long as the boot environment and the integrity of the
hardware itself are protected, helps prevent the acquisition of data. Traditional methods of encryption, however, do not provide
significant protection of a device while it is running, particularly if the attacker has a way of bypassing the lock screen.

This thesis has proposed a method of encryption capable of protecting a device even if it is seized while running.  The key
distinction between the method proposed - eCryptfs boundary mode - and traditional mobile full-disk encryption, is that eCryptfs
boundary mode independently keys Android applications. The keys of select applications can be wiped whenever the device is locked,
leaving the rest of the data accessible. This completely prevents logical acquisition of the chosen applications, and could be
extended to thwart live memory analysis. Coupling this technique with other projects that provide secure communications, such as
Orbot and TextSecure, would provide a privacy-oriented mobile OS capable of withstanding significant forensic inquiry. Properly
used, such a platform would afford individuals a much greater level of confidence that, even if they lose control of their device,
their data is safe.
